Product plots

### 2011-06-20 reviews/response.txt

# Summary review

The reviewers judge Product Plots to be a useful and parsimonious model
for constructing a large variety of plots and potentially also an example
that might spur further research in formal models of visualization
construction. All reviewers agree that this work makes a substantive
contribution to InfoVis research and argue for acceptance.

That said, reviewers also raise a number of issues that the authors
should address prior to publication. Most notable among these is the
treatment of related work. R1 and R4 in particular list elements of prior
work that should be considered and discussed. I recommend that the
authors introduce a dedicated related work section in their next
revision.

In addition, the reviewers make a number of helpful suggestions that
might inform future drafts. I strongly encourage the authors to revise
the paper accordingly.

# Major changes

* Related work section & missing references
* Improve justification of rectangles using references provided.

# Review 2

* First, related work is distributed throughout the paper. A consolidated
related work section addressing similar work and clearly delineating the
contributions of the present work is needed. Other visualization models
of interest include Slingsby et al's HiVE, Wilkinson's Grammar of
Graphics, Mackinlay's APT, and Stolte et al's Polaris. A detailed
comparison with such prior work is a pre-requisite to publication. How do
such models compare in terms of expressiveness and notational efficiency?

* I'm curious how much effort is necessary to use product plots in
practice. Some examples of statements in R and the corresponding plots
should suffice (and potentially examples of an interactive UI for product
plot construction).

* Is anyone using the system to perform analyses? While not strictly
necessary, example analysis case studies supported by product plots might
help convey the value of the model.

* "the areas of rectangles are easier to compare than areas of other shapes
[5-7]" -> This is somewhat glibly stated. A quick review of the cited
papers does not support the claim: [5] does not assess area empirically;
[6] assesses circular and arbitrary polygonal areas, but not rectangles;
[7] compares circles and rectangles, but the difference does not appear
to be statistically significant. Later in the paper, there is a similar
critique of non-rectangular polygons using the results of [6]. I don't
think this is appropriate -- [6] uses irregular polygons, whereas
techniques like voronoi treemaps generate convex polygons with average
bounding box aspect ratios near unity. Thus the argument here is also
unsubstantiated. It may well be that rectangles are better, but this has
not been convincingly demonstrated. I recommend the authors consider
dropping these perceptual arguments.

# Review 4

* A surprising omission is the work of Wilkinson (1999, 2005) on The
 Grammar of Graphics. Wilkinson introduced the idea of combining graphical
 elements through products of sets (called facets) and embedding these in
 different coordinate systems. In fact, Figure 12 echoes several of the
 figures in the Coordinates chapter of that book. Wilkinson was also the
 first to draw a link between the recursive partitioning in treemaps and
 mosaics and a set product operator. As he explains in Chapter 16 of the
 second edition, the algebra for specifying visualizations of product
 models maps one-to-one to the most common factorial experimental design
 algebra for log-linear models or ANOVA. In any case, the framework in
 this paper appears closer to that of Wilkinson's than to any of the other
 papers cited.

 Bottom of page:
 Using the word "all" is a bit strong here. Better to delete the word and
 confine the discussion to "area" plots. There are exceptions, as the
 authors point out later, so avoid the inclusive term.

 The claim for rectangles at the bottom of page 1 is not supported by
 research. The authors, like many statisticians and visualization
 researchers, tend to cite statisticians and computer scientists for
 supportive psychological principles; however, the perceptual and
 cognitive research by non-psychologists tends to have serious
 methodological flaws. I've provided some relevant citations below. There
 are significant biases in area judgments that are exacerbated by extreme
 aspect ratios (something even Kong, Heer & Agrawala noticed in Treemaps).
 The perceptual combining rule for rectangles appears to be more Attneave
 (absolute metric) than product (multiplicative), although see Algom et
 al. for support for the multiplicative rule. The authors need to rephrase
 this statement, eliminating the word "must."

 Also, Simken and Hastie, in one of the few perception articles published
 by psychologists in a statistical journal, demonstrated conclusively that
 pie charts outperform rectangular area charts for proportion of whole
 data. There are some exceptions to their findings, but statisticians need
 to understand that they are largely wrong in their scorn for pie charts.

 We can tile triangles with triangles. And these are "simpler" than
 rectangles. One can also tile the sphere with triangles, which cannot be
 done with rectangles. You need to rethink this assertion.

 References:
 Algom, D., Wolf, Y., & Bergman, B. (1985). Integration of stimulus
 dimensions in perception and memory: Composition rules and psychophysical
 relations. Journal of Experimental Psychology: General, 114, 451-471.

 Anderson, N.H. and Weiss, D.J. (1971). Test of a multiplying model for
 estimated are of rectangles. American Journal of Psychology, 84, 4,
 543-548.

 Borg, I. and Leutner, D. (1983). Dimensional models for the perception of
 rectangles. Perception & Psychophysics, 34, 257-267.

 Kidman, G. (1997). Area integration rules. In F. Biddulph & K. Carr
 (Eds.), People in mathematics (Proceedings of the 20th annual conference
 of the Mathematics Education Research Group of Australasia (Vol. 1, pp.
 271 - 277).

 Mates, J., Di Maio, V., and Lansky, P. (1992). A model of the perception
 of area. Spatial Vision, 6, 101-116.

 Morgan M.J. (2005). The visual computation of 2-D area by human
 observers. Vision Research, 45(19), 2564-70.

# Review 1
 Finally, I wonder if the authors considered any other graphical
 composition method than nesting areas? While nesting areas allowed them
 to encompass a number of existing plot types, other composition methods
 might be more useful for extension to higher dimensions. For example,
 what about simply showing a separate plot for each component?

 -label all appropriate plots in the paper with the corresponding
 distribution decomposition (directly label, not just mention in the text
 or in the caption)
 -flesh out section 3.1. Assuming that infovis readers will not know what
 conditional and marginal distributions are, the authors should
 demonstrate how they are computed from the joint distribution. Providing
 the graphical analogs to the tables in Figure 4 would be extremely
 helpful. (I realize that a marginal and a conditional distribution from
 Figure 4 are represented in Figure 5, but the connection is never made
 explicitly in the text. Some hand holding would be very helpful here.)
 Hand hold better the analytic questions that each distribution helps
 answer. E.g. "What proportion of males are very happy? The bottom left
 entries tells us it's 31%."

 Minor comments----

 Why squarified treemaps? Heer work showing non-square treemaps are easier
 to read.

 Section 3.1

 Its not clear from the wording in the first paragraph that the marginal
 and conditional distributions contain less information than the joint
 distribution. If you're going to assume that your readers don't know what
 the marginal and conditional distributions are (plausible for the InfoVis
 crowd), then it would make sense to briefly describe how they are
 computed from the joint distribution.

 Fix the formatting of the tables in Figure 4.

 Clarify in the last paragraph that in addition to the 3 factorings,
 permutations of the variables leads to more variations.

 "It remains to find an [sic] graphical analog of this multiplication."
 seems unnecessarily dramatic. We already know you found one.

 Section 3.2

 Label Figure 5 with the distribution forms, f(happy), f(happy|sex),
 f(happy, sex), rather than just mention them in the text.

 Figure 7 shows, f(marital | age) f(age) and f(age | marital) f(marital),
 right? Say this in the text and label figures.

 "the relationship between proportion and area is only constant within a
 level" -- is there a better way to say this? I think you mean level as in
 nesting level, not level as is a level of a factor, right? In Figure 7
 (right), the nested bar charts are all sized proportionally with respect
 to each other even though they occur in different levels of the marital
 status factor.

 Figure 8 middle would be better with a bar chart, right? Why show both
 the middle and right plots; they're basically the same.

 Section 6.2

 Why limit yourself to a finite number of bins? Push the number of bins to
 infinity and you can display line graphs (density plots) with hbar,
 stacked line graphs (hbar+vspine), relative area (hspine+vspine),
 overlaid line charts (hbar+hbar), or (hspine+hbar) (don't know the name
 of this chart, basically overlaid line charts, but where the sum of
 values at each vertical slice is 1).

# Review 3

 I had some problems, though, in classifying this paper - it falls
 somewhere between theory/model (describes a theoretical framework that
 encompasses a large number of existing visualisations) and technique
 (code is provided to implement the framework). I'm not sure if it'd be
 better to simplify the classification by omitting mention of the code
 except as a note near the end - while it's interesting, it's not core to
 the paper.

 Doing this would leave more space to address some other minor concerns
 with the figures: in figures 2, 3, 5, 6 and 7 it's quite clear that the
 labels on the plot have spacing issues that could be resolved with larger
 figures. Additionally, figure 4 spills over the gutter into the text in
 the right column.

 In terms of text, I noticed some typos (first page, "This framework is
 based on the fundamental statistical idea that high any high dimensional
 distribution" has too many "high"s), one slight miss in phrasing ('weed
 down' sounds like a mixed metaphor between 'weed out' and 'cut down') and
 perhaps the last column of the conclusion could benefit from editing
 (repetition of 'generally' stands out).

# Minor changes

* There is a line break after "Abstract-". The abstract text should begin
on the same line.

* Figure 4 overruns its column.

* Page 2. "Uniformly distributed" suggest to me a uniform random distribution. Is there another way to say this?

* Figure 3. Left and Right are Top and Bottom. LaTeX refused to obey your
orders, as usual ;o)

* Please italicize n in the fourth line of Section 3. Check that other
parameters are properly italicized in the text.

* Page 1. "This framework is based on the fundamental statistical idea that
 high any high dimensional" Delete the first "high."

 "51 020 and 5 000" use comma delimiters.

*  Figure 12. The bottom row is not produced by transforming after rotation.
 It is produced by switching the assignment of x and y to r and theta.
 This is a pivoting operation.

### 2011-06-21 prodplots-infovis.tex
Check in a few final tweaks

We propose a new framework for visualising tables counts, proportions or probabilities. We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions. The framework, with extensions, is sufficient to encompass over 20 visualisations previously described in fields of statistical graphics and infovis, including bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams.

@@

We begin in Section~\ref{sec:primitives} by motivating the three constraints that give rise to the basic graphical component of product plots, the rectangle. We constrain the rectangle further to produce three 1d atoms (bar, spine, tile) and one 2d atom (the fluct). These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include more visualisations such as histograms, pie charts, cascaded treemaps and weighted plots.

Section~\ref{sec:combination} provides the mathematical framework that allows us to combine the 1d and 2d atoms to display data of any dimensionality. This framework is based on the fundamental statistical idea that high any high dimensional distribution can be factorised into a product of low-dimensional conditional and marginal distributions. Section~\ref{sec:existing} shows how many existing named graphics are special cases of this general pattern.

@@

To illustrate these ideas, we will use the same data in all examples. The data is a small sample of variables related to happiness from the general social survey ({\sc gss}) \citep{davis:2008}. The {\sc gss} is a yearly cross-sectional survey of Americans, run from 1976. We combine data for 25 years to yield 51\,020 observations, and of the over 5\,000 variables, we select nine related to happiness, as described in Table~\ref{tbl:happy}.

@@

  \item Area must be {\bf proportional} to count. This is the key constraint underlying all area plots. The total area for a graphic is usually constrained, which means that area plots display typically proportions or probabilities, rather than counts. We will use count, probability and proportion interchangeably throughout the paper.

@@

This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s, a computationally trivial operation. It remains to find an graphical analog of this multiplication.

@@

We connect probability products to our rectangular primitives by noting that areas are also products, products of height and width. It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple 1d primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart. From left to right, we have plots of $f(happy)$, $f(happy | sex)$ and the product $f(happy, sex)$. The heights and widths of the rectangles are multiplied in the same way as the components of the {\sc pmf}.

@@

Formally, a product plot is constructed by the following recursive partitioning algorithm, which takes three parameters: {\sf data}, a multi-dimensional array, with dimensions ordered in the same way as the desired partitions; {\sf bounds}, a vector giving the top, left, right and bottom boundaries; and {\sf divider}, a list of the desired low-d drawing primitives. The following algorithm gives an rough idea of the computation:

@@

Different partitions reveal different features of the data. Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}. Instead of visualising the joint distribution with a fluct, we could focus on the conditional distribution of marital status given age, or age given marital status. Figure~\ref{fig:marital} shows two ways to do this. The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plot $f(age, marital)$ with a hbar nested in a vspine. These displays show the same data, but support different comparisons: on the left, we can see that most young people are unmarried, and on the right, that few unmarried people are over the age of 30. The right plot, with bars nested inside spines, also illustrates an important feature of non-space filling tilings: the relationship between proportion and area is only constant within a level. The bars are scaled to be as tall as possible, without overflowing any bounding region.

@@

Conditioning is also an important tool by itself, because it allows one to remove relationships that are known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because we must compare areas, not positions. Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness given sex and health, because comparing positions along a common scale is an easier perceptual task. Depending on the comparison we are most interested in, we can make it easier to compare across wealth given health, or health given wealth. Here we see that for a fixed income level, better health is correlated to increased happiness. The same is not true for a fixed level of health: rich people with poor health seem to be less happy than poorer people in poor health.

@@

Labelling product plots is challenging. In this paper, we use a carefully selected combination of colour and axis labels, as well displaying the process by which a product plot is created. Axis labels are only available under certain conditions: when we have rows or columns that all share the same dimensional index. This can occur because of the structure of the graphical primitive (e.g. bars, flucts), or because the display is of a conditional distribution. Take Figure~\ref{fig:marital} as an example: the left plot can only have the $x$ axis labelled, but the right plot can have both axes labelled. There are other ways to label the cells apart from colour. \citep{slingsby:2009} uses text labels, sized to occupy the space of the region they label. In the interactive setting, dynamic labelling in the form of tool tips is extremely helpful, as is the ability to use linked brushing to connect high-d plots to low-d plots.

@@

We have assumed that the proportions represent counts, but without loss of generality, we can use any set of non-negative, additive weights. For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights. These are use to account for oversampling of black respondents in certain years, and to reduce the effect of non-response by other demographics. Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex. The distribution is barely different, and suggests that we don't need to worry about weights for this plot. (Unfortunately for this dataset of we have been unable to find any plots where weighting makes a big difference.)  In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities. Some examples of weighted data graphics can be found in \citep{unwin:1998,unwin:2003aa,unwin:2006}.

@@

The framework can be trivially extended to work with continuous data: just  bin continuous variables to make them discrete. There are many different to ways to create intervals for continuous data, but two are most important \citep{denby:2009}: intervals of equal width, and intervals containing equal numbers of points. This extension allows the product plot framework to also describe {\bf histograms} and {\bf spinograms} \citep{hummel:1996}, continuous analogues of the bar and charts, shown in Figure \ref{fig:cont-examples}. A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data.

@@

With one more extension, displaying the innermost proportion with colour (instead of area), we can also describe {\bf dimensional stacking} \citep{leblanc:1990} as a plot of binned, conditioned continuous data, with proportion shown by colour.

@@

It can be useful to violate the constraint that area is proportional to value to distinguish between zeros, missing values and very small values. A zero weight should have zero area, but giving it positive area can be useful that it is visible. In general, it's useful to constrain all areas to be above a certain minimal perceptible size (e.g.\ 4 square pixels). Areas which are constrained in such a way need a visual flag (such as a different colour) to ensure that the reader knows that the relationship between area and value has been violated. This type of non-linear mapping has been implemented in {\sc manet} \citep{unwin:1996} and in hierarchical pixel bar charts \citep{keim:2002}. At the other end of the spectrum, it can be useful to constrain the size of largest values to get censored zooming \citetext{Antony Unwin, priv. comm.}, which makes it possible to focus on small values.

@@

Cascaded treemaps \citep{lu:2008} illustrate how the violation of containment can be productive. In the {\bf cascaded treemap}, each level is slightly offset from the one above to create a pseudo-3d perspective. This makes it easier to see all the levels of the hierarchy, not just the lowest level. Figure~\ref{fig:cascade} shows an example of how cascading can help illuminate the structure of a complex mosaic plot. This technique is probably more effective when implemented interactively.

@@

A pie chart is a popular method of displaying proportions, but it is not a rectangular partition and so does not seem to fall in the framework of this paper. However, there is a simple relationship between product plots and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius. Many other circular displays turn out to be special cases of product plots drawn in polar coordinates \citep{draper:2009}. We have identified the following radial plots as polar transformations of product plots:

@@

@@

The product plots framework is successful at describing many existing graphics for tables of counts, proportions and probabilities. It lays the framework for much future work, particularly because the framework leads to a combinatorial explosion of possibilities. For example, a 4d {\sc pmf}, $f(a, b, c, d)$, can be factorised in five different ways:

@@

\noindent There are 24 possible ways of ordering the variables in the {\sc pmf}, 5 ways of displaying a 1d {\sc pmf}, and 1 way of displaying a 2d {\sc pmf}, leading to a possible $24 * (5^4 + 5^2 + 5^2 + 5^2 + 1) = 16,824$ plots, before we even consider conditioning! Well thought out interaction can make it easy to navigate this space. We have a prototype interaction model developed as part of the cranvas suite of interactive graphics in R, {\tt http://github.com/ggobi/cranvas}, but much work and user testing remains to be done.

We also are hopeful that developing the connection between product plots and log-linear (aka Poisson) models will also help better understand this space of plots. Some special cases have already been worked out. For example, \citep{hofmann:2003} shows how looking for straight lines in a mosaic plot corresponds to a formal test of independence between two variables. Generally, what visual comparisons are equivalent to what formal statistical tests? And conversely, how can inspecting the significant coefficients in a model help us choose a set of useful plots for a given dataset? Generally, given a question, can we suggest appropriate plots? Or given a plot, can we suggest questions that it might answer?

Product plots need to be extended with systematic ways of displaying uncertainty, to help users identify whether differences in the plot represent real differences in the underlying population or are just the result of random variation. This is particularly important because area plots typically have fixed area, meaning that the total number of observations in a plot is not directly displayed. This is crucial for determining whether a difference is truly significant or not.

>>>>

We propose a new framework for visualising tables of counts, proportions and probabilities. We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions. The framework, with extensions, is sufficient to encompass over 20 visualisations previously described in fields of statistical graphics and infovis, including bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams.

@@

We begin in Section~\ref{sec:primitives} by motivating the three constraints that give rise to the basic graphical component of product plots: the rectangle. We constrain the rectangle further to produce three 1d atoms (bar, spine, tile) and one 2d atom (the fluct). These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include more visualisations such as histograms, pie charts, cascaded treemaps and weighted plots.

Section~\ref{sec:combination} provides the mathematical framework that allows us to combine the 1d and 2d atoms to display data of any dimensionality. This framework is based on the fundamental statistical idea that any high dimensional distribution can be factorised into a product of low-dimensional conditional and marginal distributions. Section~\ref{sec:existing} shows how many existing named graphics are special cases of this general pattern.

@@

To illustrate these ideas, we will use the same data in all examples. The data is a small sample of variables related to happiness from the general social survey ({\sc gss}) \citep{davis:2008}. The {\sc gss} is a yearly cross-sectional survey of Americans, run from 1976. We combine data for 25 years to yield 51,020 observations, and of the over 5,000 variables, we select nine related to happiness, as described in Table~\ref{tbl:happy}.

@@

  \item Area must be {\bf proportional} to count. This is the key constraint underlying all area plots. The total area for a graphic is usually constrained, which means that area plots display typically proportions or probabilities, rather than counts. With this proviso, we will use count, probability and proportion interchangeably throughout the paper.

@@

This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s, a computationally trivial operation. It remains to find a graphical analog of this multiplication.

@@

We connect probability products to our rectangular primitives by noting that areas are also products: products of height and width. It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple 1d primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart. From left to right, we have plots of $f(happy)$, $f(sex | happy)$ and the product $f(happy, sex)$. The heights and widths of the rectangles are multiplied in the same way as the components of the {\sc pmf}.

@@

More precisely, a product plot is constructed by the following recursive partitioning algorithm, which takes three parameters: {\sf data}, a multi-dimensional array, with dimensions ordered in the same way as the desired partitions; {\sf bounds}, a vector giving the top, left, right and bottom boundaries; and {\sf divider}, a list of the desired low-d drawing primitives. The following algorithm gives an rough idea of the computation:

@@

Different partitions reveal different features of the data. Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}. Instead of visualising the joint distribution with a fluct, we could focus on the conditional distribution of marital status given age, or age given marital status. Figure~\ref{fig:marital} shows two ways to do this. The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plots $f(age, marital)$ with a hbar nested in a vspine. These displays show the same data, but support different comparisons: on the left, we can see that most young people are unmarried, and on the right, that few unmarried people are over the age of 30. The right plot, with bars nested inside spines, also illustrates an important feature of non-space filling tilings: the relationship between proportion and area is only constant within a level. The bars are scaled to be as tall as possible, without overflowing any bounding region.

@@

Conditioning is also an important tool by itself, because it allows one to remove relationships that are known or uninteresting. Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. The left plot displays raw proportions, showing that most people are in good health and average financial standing. However, it is difficult to see how happiness varies within these conditions because we must compare areas, not positions. Conditioning on financial status and health produces the plots on the right (equal bin size plots) and makes it easier to see the conditional distribution of happiness given sex and health, because comparing positions along a common scale is an easier perceptual task. Depending on the comparison we are most interested in, we can make it easier to compare across wealth given health, or health given wealth. Here we see that for a fixed income level, better health is correlated to increased happiness. The same is not true for a fixed level of health: rich people with poor health seem to be less happy than poorer people in poor health.

@@

Labelling product plots is challenging. In this paper, we use a carefully selected combination of colour and axis labels, as well displaying the process by which a product plot is created. Axis labels are only available under certain conditions: when we have rows or columns that all share the same dimensional index. This can occur because of the structure of the graphical primitive (e.g. bars, flucts), or because the display is of a conditional distribution. Take Figure~\ref{fig:marital} for example: the left plot can only have the $x$ axis labelled, but the right plot can have both axes labelled. There are other ways to label the cells apart from colour. \citep{slingsby:2009} uses text labels, sized to occupy the space of the region they label. In the interactive setting, dynamic labelling in the form of tool tips is extremely helpful, as is the ability to use linked brushing to connect high-d plots to low-d plots.

@@

We have assumed that the proportions represent counts, but without loss of generality, we can use any set of non-negative, additive weights. For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights. These are used to account for oversampling of black respondents in certain years, and to reduce the effect of non-response in other demographics. Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex. The distribution is barely different, and suggests that we don't need to worry about weights for this plot. (Unfortunately for this dataset of we have been unable to find any plots where weighting does makes a difference.)  In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities. Some examples of weighted data graphics can be found in \citep{unwin:1998,unwin:2003aa,unwin:2006}.

@@

The framework can be trivially extended to work with continuous data: just  bin continuous variables to make them discrete. There are many different to ways to create bins for continuous data, but two are most important \citep{denby:2009}: bins of equal width, and bins containing equal numbers of points. This extension allows the product plot framework to also describe {\bf histograms} and {\bf spinograms} \citep{hummel:1996}, continuous analogues of bar and spine charts, shown in Figure \ref{fig:cont-examples}. A long standing tradition is that no gaps are displayed between adjacent rectangles when displaying continuous data.

A more theoretical approach in dealing with continuous variables  lies in increasing the number of bins infinitesimally -- this leads from a probability mass function to a density function, making the 1d primitives of hbar and vbar to density plots, and changing the spine-plot representation to a conditional density plot (citation needed - probably Martin's and Heike's paper). While the inclusion of one continuous variable is completely feasible that way, two variables need the use of an additional aesthetic, such a s colour,  to visualize a 2d joint density. Three or more continuous variables break the scope of the product plots framework.

@@

With one more extension, displaying the innermost proportion with colour (instead of area), we can also describe {\bf dimensional stacking} \citep{leblanc:1990}.

@@

It can be useful to violate the constraint that area is proportional to value to distinguish between zeros, missing values and very small values. A zero weight should have zero area, but giving it positive area can be useful so that it is visible. In general, it's useful to constrain all areas to be above a certain minimal perceptible size (e.g.\ 4 square pixels). Areas which are constrained in such a way need a visual flag (such as a different colour) to ensure that the reader knows that the relationship between area and value has been violated. This type of non-linear mapping has been implemented in {\sc manet} \citep{unwin:1996} and in hierarchical pixel bar charts \citep{keim:2002}. At the other end of the spectrum, it can be useful to constrain the size of largest values to get censored zooming \citetext{Antony Unwin, priv. comm.}, which makes it possible to focus on small values.

@@

Cascaded treemaps \citep{lu:2008} illustrate how the violation of containment can be productive. In the {\bf cascaded treemap}, each level is slightly offset from the one above to create a pseudo-3d perspective. This makes it easier to see all the levels of the hierarchy, not just the lowest level. Figure~\ref{fig:cascade} shows an example of how cascading can help illuminate the structure of a complex mosaic plot. This technique is probably most effective when implemented interactively.

@@

A pie chart is a popular method of displaying proportions, but it is not a rectangular partition and so does not seem to fall in the framework of this paper. However, there is a simple relationship between product plots and pie charts: a pie chart is an hspine drawn in polar coordinates with the $x$ coordinate mapped to angle and the $y$ coordinate to radius. Many other circular displays turn out to be special cases of product plots drawn in polar coordinates \citep{draper:2009}. We have identified the following radial plots as polar transformations of product plots:

@@

\section{Related Work}
\label{sec:related}

The product plots framework is heavily influenced by the work of WIlkinson \citep{wilkinson:1999, wilkinson:2005}, who introduced the set product operator to map hierarchical data to a two-dimensional framework.  Wilkinson restricts (??? not quite sure - need to lookit up) the use to standard mosaicplots as defined by \citet{hartigan:1981}, which could easily be extended, however the product plots framework additionally introduces the concept of conditioning on variables.

 \citep{slingsby:2009}: the HiVE system provides a notation allowing to describe states of hierarchical layouts and operators for reconfiguration. HiVE allows both rectangular layouts and polygon layouts -- out of the rectangular layouts, all but the spatially ordered algorithm are covered by the  framework of product plots.

The framework of  Polaris  \cite{stolte:2002} relies heavily on a tabular or cubic format in rendering hierarchical graphics, which makes these charts essentially a three dimensional extension of the trellis framework \citep{cleveland:1994, cleveland:1996}:  in the product plots framework, this approach is captured by conditioning on the variables used in rows and columns, and displaying the other variables unconditioned in each cell of the resulting tabular layout.

 \cite{mackinlay:1986, mackinlay2007}: VisQL, a visual query language is designed to capture the flow of a graphical data exploration.

@@

The product plots framework is successful at describing many existing graphics that display tables of counts, proportions and probabilities. It lays the framework for much future work, particularly because the framework leads to a combinatorial explosion of possibilities. For example, a 4d {\sc pmf}, $f(a, b, c, d)$, can be factorised in five different ways:

@@

\noindent There are 24 possible ways of ordering the variables in the {\sc pmf}, 5 ways of displaying a 1d {\sc pmf}, and 1 way of displaying a 2d {\sc pmf}, leading to a possible $24 * (5^4 + 5^2 + 5^2 + 5^2 + 1) = 16,824$ plots, before we even consider conditioning! The product plots framework defines a large space of potential plots.

Well thought out interaction will make it easier to navigate this space, and we have begun to develop a prototype model in R, as part of the cranvas suite of interactive graphics, {\tt http://github.com/ggobi/cranvas}.  However, much work and user testing remains to be done before we can be confident that we have developed a useful navigation model.

We are also exploring the connection between product plots and log-linear (aka Poisson) models, the statistical models most commonly used for count data. Some special cases of the general connection have already been worked out. For example, \citep{hofmann:2003} shows how looking for straight lines in a mosaic plot corresponds to a formal test of independence between two variables. Generally, we are interested in knowing what visual comparisons are equivalent to what formal statistical tests, and conversely, how significant coefficients in a model can help us choose a set of useful plots. Generally, given a question, can we suggest appropriate plots? Or given a plot, can we suggest questions that it might answer?

Product plots also need to be extended with systematic ways of displaying uncertainty, to help users identify whether differences in the plot represent real differences in the underlying population or are just the result of random variation. This is particularly important because area plots typically have fixed area, meaning that the total number of observations in a plot is not directly displayed, even though this is crucial for determining whether a difference is statistically significant or not.
